---
title: "Introduction to algebra" 
author: "Erika Duan"
date: last-modified
format: 
    gfm:
        html-math-method: webtex   
        toc: true
        toc-depth: 2   
        toc-title: Contents  
execute:   
  echo: true     
  output: false   
  warning: false  
---

::: callout-tip
## Summary

This tutorial lists some common algebraic tricks, such as factorisation and working with inequalities, which are useful for solving mathematical problems in later tutorials.
:::

```{r}
#| echo: false
#| output: false

# Install R packages -----------------------------------------------------------
# The native pipe operator requires R version 4.1+ 
packages <- c("ggplot2",
              "dplyr")

installed_packages <- packages %in% rownames(installed.packages())

if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Attach R packages to the global environment
library("dplyr")
library("ggplot2")
```


# Elementary algebra

Elementary algebra is commonly associated with questions like *Solve for x when $2x + y = 6$.* It is more useful, however, to think of elementary algebra as a form of expression that allows us to represent infinitely possible terms using finite ones.

For example, the terms $6x + 3y$, $4x + 2y$ and $2x + y$ are equivalent and there are theoretically infinite variations of this expression, with the simplest being $2x + y$. When we read the term $2x + y$, we can intuit that 2 parts of $x$ and 1 part of $y$ are always required to make up a final quantity.

![](../figures/algebra-representation.svg)

An algebraic term can therefore be decomposed into three parts:

-   A single **variable** or multiple variables. A variable is a varying quantity of an entity, usually represented by concise symbols $x$, $y$ or $x_i$ where $i=1, \cdots, n$.
-   The relative **quantity** of each variable. *For example, let $x$ represent the number of eggs and $y$ represent the number of cups of sugar required to make a dessert.* The term $x + y$ describes a 1:1 ratio of eggs to sugar whereas the term $2x + y$ describes a 2:1 ratio of eggs to sugar and will result in a very different taste.
-   The arithmetic **operator(s)** applied between variables. *For example, in additive models, the relationship between parameters $y = b_o+b_1x_1 + \cdots + b_nx_n$ is additive and the dependent variable $y$ therefore increases by $b_1$ amount for one unit increase in $x_1$.*


# Manipulating inequalities

Algebraic terms using inequalities are common when we want to prove the existence of an upper or lower bound. *For example, if A is an event in the probability space, we know that the probability of event A occurring is between 0 and 1 inclusive i.e.* $0 \leq P(A) \leq 1$.

There are three rules for manipulating inequalities:

-   Adding or subtracting the same quantity from both sides of an inequality leaves the inequality symbol unchanged.
-   Multiplying or dividing both sides of an inequality by a positive number leaves the inequality symbol unchanged.
-   Multiplying or dividing both sides of an inequality by a negative number reverses the inequality symbol.

![](../figures/algebra-inequality_rules.svg) 

::: panel-tabset
## R

In R, inequality statements are outputted as Boolean values i.e. `TRUE` or `FALSE`.  

```{r}
# Compute inequality in R ------------------------------------------------------
4 <= 4
#> [1] TRUE

4 < 4
#> [1] FALSE

class(4 < 4)
#> [1] "logical"
```

## Python

In Python, inequality statements are also outputted as Boolean values i.e. `True` or `False`.

```{python}
# Compute inequality in Python -------------------------------------------------
4 <= 4
#> True 

type(4 <= 4)
#> <class 'bool'>  
```

## Julia

In Julia, inequality statements are also outputted as Boolean values i.e. `true` or `false`.

```{julia}
# Compute inequality in Julia --------------------------------------------------
4 <= 4
#> true 

typeof(4 <= 4) 
#> true
#> Bool

a = 1
b = 2
c = 3

a < b, a + c < b + c
#> (true, true) 
```
:::


# Simplifying fractions

# Factorisation

# Resources

-   [Entry](https://plato.stanford.edu/entries/algebra/) on algebra from the Stanford Encyclopedia of Philosophy.\
-   Khan academy [YouTube series](https://www.khanacademy.org/math/algebra-basics/alg-basics-algebraic-expressions#alg-basics-intro-to-variables) on algebra basics.\
-   A [factsheet](https://www.mathcentre.ac.uk/resources/uploaded/mc-bus-manipinequ-2009-1.pdf) on manipulating inequalities from the Uk Maths Centre.
