---
title: Introduction to functions     
author: Erika Duan
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 2
    math_method:
      engine: webtex
      url: https://latex.codecogs.com/svg.format?
---

```{r setup, include=FALSE}
# Set up global environment configuration --------------------------------------
knitr::opts_chunk$set(echo=TRUE,
                      results='hide',
                      fig.show='hold',
                      fig.align='center',
                      message=FALSE,
                      warning=FALSE,
                      out.width='80%')

knitr::knit_engines$set(python = reticulate::eng_python)
```

```{r, echo=FALSE}
# Install R packages -----------------------------------------------------------
# The native pipe operator requires R version 4.1+ 
packages <- c("ggplot2",
              "dplyr",
              "patchwork")

installed_packages <- packages %in% rownames(installed.packages())

if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Attach R packages to the global environment
library("dplyr")
library("ggplot2")
library("patchwork")

# Check version of Python used by reticulate -----------------------------------
reticulate::py_config()
```


# Properties of functions  

A **function** can be thought of as a rule that maps each element $x$ in set A to exactly one element $y=f(x)$ in set B.    

+ The domain of $f: A \to B$ is the set A, which is usually $\mathbb{R}$ or $\mathbb{R}^n$.  
+ The domain can be restricted by the function form. For $f(x) = \sqrt{x_2-4}$, the domain is $\{x \in \mathbb{R} | x \geq 2$ or $x \leq -2 \}$ or $(- \infty, -2] \cup [2, \infty )$.    
+ The co-domain of $f: A \to B$ is the set B, which is usually $\mathbb{R}$ or $\mathbb{R}^n$.   
+ The range of $f: A \to B$ is $R(f) = \{y | y = f(x)$ for some $x \in A\}$. 
+ The range is more restricted than the co-domain. For $f(x) = sin(x)$, the co-domain is $\mathbb{R}$ and the range is $[-1, 1]$.   
    
```{r, echo=FALSE, results='markup'}
knitr::include_graphics("../figures/functions-simple_function_example.svg")
```

For $f(x)$ and $g(x)$, if the range of $g(x)$ is in the domain of $f(x)$, the **function composite** $(f\circ g)(x) = f(g(x))$ exists. For example, if $f(x) = log(x)$ and $g(x) = \sqrt{x+1}$, then $(f\circ g)(x)=log(\sqrt{x+1})$ and $(g\circ f)(x) = \sqrt{log(x)+1}$.    


# Finding the function limit  

The limit of a function $lim_{x \to a} f(x) = L$ comprises values of $f(x)$ for values of $x$ close to but not equal to $a$. This is equivalent to saying that $f(x)$ approaches the limit $L$ as $x$ approaches $a$.    

The **formal definition** of a function limit involves:   

+ Choose $\epsilon$ and $\delta$ such that $\epsilon > 0$ and $\delta > 0$.    
+ Solve $0 < |x-a| < \delta$.   
+ Confirm that when $x \neq a$, $|f(x) - L| < \epsilon$.    

This method is not used in practice as it is very laborious to find $\delta$. For example, in order to verify that $lim_{x \to a} x^2 = 4$, we need to:     

+ Factorise $|x^2 - 4| < \epsilon$ into $|x - 2||x + 2| < \epsilon$.  
+ Manually calculate what happens to the $|x + 2|$ when $|x - 2| < 1$. When $|x - 2| < 1$, $1 < x < 3$ and therefore $3 < x + 2 < 5$.   
+ Substitute $3 < x + 2 < 5$ into $|x - 2||x + 2| < \epsilon$. Therefore $|x - 2||x + 2| < 5|x - 2| < \epsilon$.  
+ Therefore, when $5|x - 2| < \epsilon$, $|x + 2| < 1$ and $|x + 2| < \tfrac{\epsilon}{5}$.  
+ Therefore $\delta = min(1, \tfrac{\epsilon}{5})$.   

As a matter of pragmatism, to find the function limit, we simplify the function or composite function using the limit laws.   

```{r, echo=FALSE, results='markup'}
knitr::include_graphics("../figures/functions-limit_laws.svg")
```

**Note:** For non-arbitrary or non-split functions, a consequence of the limit laws is that $lim_{x \to a} f(x) = f(a)$ when $a$ is in the domain of $f(x)$.    

Types of limits:  

+ Left limit: if $f(x)$ is defined for all $x < a$ where $x$ is near $a$.    
+ Right limit: if $f(x)$ is defined for all $x > a$ where $x$ is near $a$.   
+ Limit at infinity: if $f(x)$ is defined for all sufficiently large positive values of $x$ and $f(x)$ is close enough to L.     
+ Limit at negative infinity: if $f(x)$ is defined for all sufficiently small positive values of $x$ and $f(x)$ is close enough to L. 

```{r, echo=FALSE, results='markup'}
knitr::include_graphics("../figures/functions-limit_types.svg")
```

**Note:** Not all functions have limits. For $lim_{x \to \infty} (x+1)$, the limit is $\infty$ and is therefore undefined. A function only has a limit if the left limit and right limit are defined and equal to each other.     

<details><summary>R code</summary>  
<p>   

```{r}
# Plot f(x) = x^2 --------------------------------------------------------------
set.seed(111)

p1 <- ggplot(data.frame(x = runif(1000, -20, 20)), aes(x)) +
  geom_function(fun = ~ (.x)^2) +
  xlim(-4, 4) + 
  ylim(-1, 4) + 
  labs(title = "f(x) = x^2") +  
  theme_minimal() + 
  theme(panel.border = element_rect(fill = NA),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linetype = "dotted")) +
  annotate("text", x = 1.4, y = -0.2, label = "D: (-infinity, infinity)") +
  annotate("text", x = 1.4, y = -0.5, label = "R: [0, infinity)") +
  annotate("text", x = 1.4, y = -0.8, label = "Undefined limit at infinity")  

# Plot f(x) = abs(x)/x ---------------------------------------------------------
set.seed(111)

p2 <- ggplot(data.frame(x = runif(1000, -20, 20)), aes(x)) +
  geom_function(fun = ~ abs(.x)/.x) +
  geom_vline(xintercept = 0, linetype = "dotted") + 
  xlim(-4, 4) + 
  ylim(-1, 4) + 
  labs(title = "f(x) = abs(x)/x") +  
  theme_minimal() + 
  theme(panel.border = element_rect(fill = NA),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linetype = "dotted")) +
  annotate("text", x = 1.8, y = -0.2, label = "D: (-infinity, infinity)") +
  annotate("text", x = 1.8, y = -0.5, label = "R: -1 or 1") +
  annotate("text", x = 1.8, y = -0.8, label = "Undefined limit at infinity")

# Plot ggplot figures side by side --------------------------------------------- 
p1 + p2
```

</p>  
</details><p>  


# The Squeeze theorem   


# Resources  
+ 