---
title: Linear transformations       
author: Erika Duan
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 2
    math_method:
      engine: webtex
      url: https://latex.codecogs.com/svg.format?
---

```{r setup, include=FALSE}
# Set up global environment configuration --------------------------------------
knitr::opts_chunk$set(echo=TRUE,
                      results='hide',
                      fig.show='hold',
                      fig.align='center',
                      message=FALSE,
                      warning=FALSE,
                      out.width='80%')

knitr::knit_engines$set(python = reticulate::eng_python)
```

```{r, echo=FALSE}
# Install R packages -----------------------------------------------------------
# The native pipe operator requires R version 4.1+ 
packages <- c("ggplot2",
              "dplyr")

installed_packages <- packages %in% rownames(installed.packages())

if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Attach R packages to the global environment
library("dplyr")
library("ggplot2")

# Check version of Python used by reticulate -----------------------------------
reticulate::py_config()
```


# Vector transformation notations       
Vectors are an important way to represent values as positions within n-th dimensional space. Vector operations and matrix-vector operations enable us to transform values in useful ways.    

A column vector $\vec u$ is an $m \times 1$ matrix where $\vec u \in \mathbb{R}^m$. Two vectors are equal if they have the same number of rows and their corresponding entries are equal.  

**Note:** Do not confuse the zero vector $\vec 0$ with the scalar value of 0 when writing vector operations.   

```{r, echo=FALSE, results='markup'}
knitr::include_graphics("../figures/linear_systems-vectors.svg")
```

# Dimensionality and linear transformations   


 
# Resources  
+ Great YouTube videos on [2D](https://www.youtube.com/watch?v=kYB8IZa5AuE&list=PLZHQObOWTQDPD3MizzM2xVFitgF8hE_ab&index=3) and [3D](https://www.youtube.com/watch?v=rHLEWRxRGiM&list=PLZHQObOWTQDPD3MizzM2xVFitgF8hE_ab&index=5) linear transformations by 3Blue1Brown.      