---
title: Introduction to vectors     
author: Erika Duan
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 2
    math_method:
      engine: webtex
      url: https://latex.codecogs.com/svg.format?
---

```{r setup, include=FALSE}
# Set up global environment configuration --------------------------------------
knitr::opts_chunk$set(echo=TRUE,
                      results='hide',
                      fig.show='hold',
                      fig.align='center',
                      message=FALSE,
                      warning=FALSE,
                      out.width='80%')

knitr::knit_engines$set(python = reticulate::eng_python)
```

```{r, echo=FALSE}
# Install R packages -----------------------------------------------------------
# The native pipe operator requires R version 4.1+ 
packages <- c("ggplot2",
              "dplyr")

installed_packages <- packages %in% rownames(installed.packages())

if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Attach R packages to the global environment
library("dplyr")
library("ggplot2")

# Check version of Python used by reticulate -----------------------------------
reticulate::py_config()
```


# Vectors   
Vectors are an important form used to convey object position in dimensional space and vector operations enable us to transform vectors in useful ways.    

A column vector $\vec u$ is an $m\times1$ matrix where $\vec u\in \mathbb{R}^m$. Two vectors are equal if they have the same number of rows and their corresponding entries are equal.  

**Note:** Do not confuse the zero vector $\vec 0$ as a scalar 0 during vector operations.  

```{r, echo=FALSE, results='markup'}
knitr::include_graphics("../figures/linear_systems-vectors.svg")
```

Key vector operations are:  

+ Scalar multiplication - when a vector is scaled by a constant.   
+ Vector addition - when two vectors of the same dimensions are added to form a new vector (with a new length and direction).   
+ Vector multiplication - used to calculate how far apart two vectors are, with respect to the origin, by superimposing one along the axis of the other.     


# Vector scalar multiplication   
Vector scalar multiplication is used to scale the length of the vector and/or to change its direction (i.e. vector direction can also be reversed by multiplication of a negative constant).   

Vector scalar multiplication is denoted as $c\vec u$ where $c\in\mathbb{R}$.  

```{r, echo=FALSE, results='markup'}
knitr::include_graphics("../figures/linear_systems-vector_scalar_multiplication.svg")  
```

**Note:** The reverse of $\vec u$ is denoted as $-\vec {u}$.   


# Vector addition   

Two vectors of the same dimension can be added entry-wise. This produces a new vector with a new length and direction in dimensional space.   

Vector addition is denoted as $\vec u + \vec v = \vec{(u+v)}$. Geometrically, in 2D space, vector addition corresponds to the 4th vertex of the parallelogram whose other vertices are $\vec 0, \vec u,$ and $\vec v$.        

```{r, echo=FALSE, results='markup'}
knitr::include_graphics("../figures/linear_systems-vector_addition.svg")  
```

**Note:** Vector subtraction is equivalent to the addition of a vector scaled by -1 i.e. $\vec u - \vec v = \vec u + (-1)(\vec v)$.   


# Vector span  



# Resources  
+ A great [YouTube series](https://www.youtube.com/watch?v=fNk_zzaMoSs) on vectors by 3Blue1Brown.      