---
title: Introduction to vectors     
author: Erika Duan
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 2
    math_method:
      engine: webtex
      url: https://latex.codecogs.com/svg.format?
---

```{r setup, include=FALSE}
# Set up global environment configuration --------------------------------------
knitr::opts_chunk$set(echo=TRUE,
                      results='hide',
                      fig.show='hold',
                      fig.align='center',
                      message=FALSE,
                      warning=FALSE)

knitr::knit_engines$set(python = reticulate::eng_python)
```

```{r, echo=FALSE}
# Install R packages -----------------------------------------------------------
# The native pipe operator requires R version 4.1+ 
packages <- c("ggplot2",
              "dplyr")

installed_packages <- packages %in% rownames(installed.packages())

if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Attach R packages to the global environment
library("dplyr")
library("ggplot2")

# Check version of Python used by reticulate -----------------------------------
reticulate::py_config()
```


# Vectors   

Vectors are an important form used to convey object position in dimensional space and vector operations enable us to transform vectors in useful ways.    
A column vector $\vec u$ is an $m\times1$ matrix where $\vec u\in \mathbb{R}^m$. Two vectors are equal if they have the same number of rows and their corresponding entries are equal.  

**Note:** Do not confuse the zero vector $\vec 0$ as a scalar 0 during vector operations.  

```{r, echo=FALSE, results='markup', out.width="90%"}
knitr::include_graphics("../figures/linear_systems-vectors.svg")
```

Key vector operations are:  

+ Scalar multiplication  
+ Vector addition  
+ Vector multiplication  

## Vector scalar multiplication  

Vector scalar multiplication is used to scale the length of the vector and/or change its direction (i.e. vector direction can also be reversed by multiplication of a negative constant).   

Vector scalar multiplication is denoted as $c\vec u$ where $c\in\mathbb{R}$.  
**Note:** The reverse of $\vec u$ is denoted as $-\vec {u}$.  

```{r, echo=FALSE, results='markup', out.width="90%"}
knitr::include_graphics("../figures/linear_systems-vector_scalar_multiplication.svg")  
```


## Vector addition   



# Resources  

+ A great [YouTube](https://www.youtube.com/watch?v=fNk_zzaMoSs) series on vectors.    
  