---
title: Linear systems     
author: Erika Duan
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 2
    math_method:
      engine: webtex
      url: https://latex.codecogs.com/svg.format?
---

```{r setup, include=FALSE}
# Set up global environment configuration --------------------------------------
knitr::opts_chunk$set(echo=TRUE,
                      results='hide',
                      fig.show='hold',
                      fig.align='center',
                      message=FALSE,
                      warning=FALSE)

knitr::knit_engines$set(python = reticulate::eng_python)
```

```{r, echo=FALSE}
# Install R packages -----------------------------------------------------------
# The native pipe operator requires R version 4.1+ 
packages <- c("ggplot2",
              "dplyr",
              "JuliaCall")

installed_packages <- packages %in% rownames(installed.packages())

if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Attach R packages to the global environment
library("dplyr")
library("ggplot2")
library("JuliaCall")

# Check version of Python used by reticulate -----------------------------------
reticulate::py_config()

# Set up Julia for automatic type conversion -----------------------------------
julia <- julia_setup()
```


# Linear systems    

There is a fundamental relationship between equations and lines. Lines represent the vector subspace of all solutions found for a linear equation.       

Depending on the dimension of the vector space, we can describe linear equations with some geometric intuition. For example, $3x_1 = 9$ is solved as $x_1 = 3$, where $x_1$ describes a point (a copy of $\mathbb{R}^0$ inside $\mathbb{R}^1$).  

```{r, echo=FALSE, results='markup', out.width="90%"}
knitr::include_graphics("../figures/numbers-categories.svg")
```



# Resources  

+ A great [Wikipedia](https://en.wikipedia.org/wiki/Number#Main_classification) explanation of the number system.     