---
title: Linear transformations       
author: Erika Duan
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 2
    math_method:
      engine: webtex
      url: https://latex.codecogs.com/svg.format?
---

```{r setup, include=FALSE}
# Set up global environment configuration --------------------------------------
knitr::opts_chunk$set(echo=TRUE,
                      results='hide',
                      fig.show='hold',
                      fig.align='center',
                      message=FALSE,
                      warning=FALSE,
                      out.width='80%')

knitr::knit_engines$set(python = reticulate::eng_python)
```

```{r, echo=FALSE}
# Install R packages -----------------------------------------------------------
# The native pipe operator requires R version 4.1+ 
packages <- c("ggplot2",
              "dplyr")

installed_packages <- packages %in% rownames(installed.packages())

if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Attach R packages to the global environment
library("dplyr")
library("ggplot2")

# Check version of Python used by reticulate -----------------------------------
reticulate::py_config()
```


# Vector transformation notation   
The essence of linear algebra involves the study of linear transformations of vector spaces.   

A linear transformation can be described as:  

+ A function which maps a vector in $\mathbb{R}^n$ to a vector in $\mathbb{R}^m$ or $\mathbb{R}^n$.    
+ This is denoted by $T: \mathbb{R}^n \to \mathbb{R}^m$ or $T: \mathbb{R}^n \to \mathbb{R}^n$ respectively.   
+ Alternatively, we can denote linear transformations with respect to its vector inputs and outputs i.e. $\vec x {\; T \;\atop \mapsto} \vec w$, where $\vec w = T(\vec x)$.    
+ The domain of $T(\vec x)$ is $\mathbb{R}^n$.   
+ The co-domain of $T(\vec x)$ can be $\mathbb{R}^m$ or $\mathbb{R}^n$ respectively.     
+ The image of $vec x$ under T is the set $\{\vec w \in \mathbb{R}^m | \vec w = T(\vec x)\}$ where $\vec x \in \mathbb{R}^n$.  
+ The range of $T(\vec x)$ also describes the set $\{\vec w \in \mathbb{R}^m | \vec w = T(\vec x)\}$ where $\vec x \in \mathbb{R}^n$.    
+ Linear transformations can also be regarded as a transformation by a standard matrix for T (draw this definition too)   

```{r, echo=FALSE, results='markup'}
knitr::include_graphics("../figures/linear_systems-linear_transformation_notation.svg")  
```



A linear transformation $T: \mathbb{R}^n \to \mathbb{R}^m$ must satisfy the following two properties:  

+ 
+ 



# Dimensionality and linear transformations   

A linear transformation $T: \mathbb{R}^n \to \mathbb{R}^m$ can be described in the following ways:  

+ 
+ 

```{r, echo=FALSE, results='markup'}
knitr::include_graphics("../figures/linear_systems-n_to_m_transformation.svg")
```

A linear transformation $T: \mathbb{R}^n \to \mathbb{R}^m$ can be described in the following ways:  

+ 
+ 

```{r, echo=FALSE, results='markup'}
# knitr::include_graphics("../figures/linear_systems-n_to_n_transformation.svg")
```


# Injective linear transformations   

A transformation is injective if: 

Every $\vec b \in \mathbb{R}^m$ is the image of at most one $\vec x in \mathbb{R}^n$. 
I.e. only one solution for $\vec b$ or the linear combination of basis vectors of A and coordinates represented by $\vec x$ only map to one solution.  

T(x) = 0 or A(x) = 0 only has the trivial solution. Therefore NulA only has the trivial subspace.  

A projection transformation is not injective.  
Different vectors in R^n always have different images in R^m.  


# Surjective linear transformations    

Every $\vec b \in \mathbb{R}^m$ is the image of one or more $\vec x in \mathbb{R}^n$. 
RangeT = Rm.  




# Matrix composition and linear transformations

 
# Resources  
+ Great YouTube videos on [2D](https://www.youtube.com/watch?v=kYB8IZa5AuE&list=PLZHQObOWTQDPD3MizzM2xVFitgF8hE_ab&index=3) and [3D](https://www.youtube.com/watch?v=rHLEWRxRGiM&list=PLZHQObOWTQDPD3MizzM2xVFitgF8hE_ab&index=5) linear transformations by 3Blue1Brown.      