---
title: Introduction to matrices     
author: Erika Duan
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 2
    math_method:
      engine: webtex
      url: https://latex.codecogs.com/svg.format?
---

```{r setup, include=FALSE}
# Set up global environment configuration --------------------------------------
knitr::opts_chunk$set(echo=TRUE,
                      results='hide',
                      fig.show='hold',
                      fig.align='center',
                      message=FALSE,
                      warning=FALSE,
                      out.width='80%')

knitr::knit_engines$set(python = reticulate::eng_python)
```

```{r, echo=FALSE}
# Install R packages -----------------------------------------------------------
# The native pipe operator requires R version 4.1+ 
packages <- c("ggplot2",
              "dplyr")

installed_packages <- packages %in% rownames(installed.packages())

if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Attach R packages to the global environment
library("dplyr")
library("ggplot2")

# Check version of Python used by reticulate -----------------------------------
reticulate::py_config()
```


# Matrices     
A matrix with $m$ rows and $n$ columns can be used to hold the coefficients from a linear system $A\vec x = \vec b$. In other words, each vector $\vec a_i$ inside matrix $A$ stores information about where the basis vector for the i-th dimension lands.  

```{r, echo=FALSE, results='markup'}
knitr::include_graphics("../figures/linear_systems-matrix_transformations.svg")
```

# Matrix column space  
For any $m\times n$ matrix A, the column space of A, denoted $ColA$, is therefore the subspace of $\mathbb{R}^m$ spanned by the columns i.e. basis vectors $\{\vec a_1, \vec a_2, \cdots, \vec a_n\}$ of matrix A.  

If we represent $A = \begin{bmatrix} \vec a_1 & \vec a_2 & \cdots & \vec a_n \end{bmatrix}$, then $ColA=Span\{\vec a_1, \vec a_2, \cdots, \vec a_n\}$.  

Therefore $ColA = \{\vec b \in \mathbb{R}^m | A\vec x = \vec b \,for \,some \, \vec x \in \mathbb{R}^n\}$.    

```{r, echo=FALSE, results='markup'}
knitr::include_graphics("../figures/linear_systems-matrix_column_space.svg")
```


# Matrix null space   
For any $m\times n$ matrix A, the null space of A is the set $NulA$ of all solutions to the homogeneous equation $A\vec x=\vec 0$.  

Therefore $NulA = \{\vec x \in \mathbb{R}^n | A\vec x = \vec 0\}$.   

The null space of A is a subspace of $\mathbb{R}^n$ as the solution set of $A\vec x = \vec 0$ can also be expressed in the parametric form $c_1\vec v_1 + c_2\vec v_2 + \cdots + c_n\vec v_h$ i.e. as $Span\{\vec v_1, \vec v_2, \cdots, \vec v_h\}$ and $Span\{\vec v_1, \vec v_2, \cdots, \vec v_h\}$ is a subspace in $\mathbb{R}^n$.  

```{r, echo=FALSE, results='markup'}
knitr::include_graphics("../figures/linear_systems-matrix_null_space.svg")
```

**Note:** $ColA$ represents the span of all basis vectors of matrix A whereas $NulA$ represents the set of all possible solutions to form matrix A when $A \vec x = \vec 0$.    

# Resources  
+ A great [YouTube series](https://www.youtube.com/watch?v=kYB8IZa5AuE&list=PLZHQObOWTQDPD3MizzM2xVFitgF8hE_ab&index=3) on matrices by 3Blue1Brown.     