---
title: Introduction to numbers   
author: Erika Duan
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 2
    math_method:
      engine: webtex
      url: https://latex.codecogs.com/svg.format?
---

```{r setup, include=FALSE}
# Set up global environment configuration --------------------------------------
knitr::opts_chunk$set(echo=TRUE, results='hide', fig.align='center')
knitr::knit_engines$set(python = reticulate::eng_python)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Install R packages -----------------------------------------------------------
# The native pipe operator requires R version 4.1+ 
packages <- c("ggplot2",
              "dplyr",
              "JuliaCall")

installed_packages <- packages %in% rownames(installed.packages())

if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Attach R packages to the global environment
library("dplyr")
library("ggplot2")
library("JuliaCall")

# Check version of Python used by reticulate -----------------------------------
reticulate::py_config()

# Set up Julia for automatic type conversion -----------------------------------
julia <- julia_setup()
```


# Numbers  

Numbers are used to:  
+ Count objects. For example, I have 1 :apple:.    
+ Describe object behaviour. For example, the area of a :white_circle: is $\pi r^2$.     
+ Solve problems, often by introducing simple assumptions. For example, if 5 :bee: can pollinate a 100 $m^2$ field in 10 minutes, how many :bee: are required to pollinate a 350 $m^2$ field? Assume that pollination efficiency scales perfectly and is not impacted by land shape.   
+ Represent complex abstractions. For example, how can individual words be represented in an n-dimensional feature space?        

Numbers can be classified into different categories according to their properties. A number can belong to multiple categories. For example, the number 2 is a count (a natural number) and a whole number (an integer), and can also be represented as a fraction i.e. $\tfrac{2}{1}$ (a rational number) and a complex number i.e. $3+i^2$.    

Number classifications have an elegant hierarchical property. All natural numbers are integers. All integers are rational numbers i.e. $n = \tfrac{n}{1}$. All rational numbers are complex numbers i.e. $\tfrac{n}{1} = \tfrac{n+1}{1}+i^2$.      


> **Note**  
> Can you think of different ways to describe the number 2 compared to the number $\tfrac{3}{2}$?    


# Natural numbers  

Natural numbers   

## Rational versus irrational numbers  

## Real numbers  

## Complex numbers  


# Resources  
+ A great [Wikipedia](https://en.wikipedia.org/wiki/Number#Main_classification) explanation of the number system.    
+ Introduction to number systems and binary numbers from [Khan Academy](https://www.youtube.com/watch?v=ku4KOFQ-bB4).   
+ Introduction to rational and irrational numbers from [Khan Academy](https://www.youtube.com/watch?v=cLP7INqs3JM).  
+ Numbers classification from [Nerdstudy](https://www.youtube.com/watch?v=vbPUS-0Wbv4).   