---
title: Introduction to numbers   
author: Erika Duan
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 2
    math_method:
      engine: webtex
      url: https://latex.codecogs.com/svg.format?
---

```{r setup, include=FALSE}
# Set up global environment configuration --------------------------------------
knitr::opts_chunk$set(echo=TRUE,
                      results='hide',
                      fig.show='hold',
                      fig.align='center',
                      message=FALSE,
                      warning=FALSE)

knitr::knit_engines$set(python = reticulate::eng_python)
```

```{r, echo=FALSE}
# Install R packages -----------------------------------------------------------
# The native pipe operator requires R version 4.1+ 
packages <- c("ggplot2",
              "dplyr",
              "JuliaCall")

installed_packages <- packages %in% rownames(installed.packages())

if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Attach R packages to the global environment
library("dplyr")
library("ggplot2")
library("JuliaCall")

# Check version of Python used by reticulate -----------------------------------
reticulate::py_config()

# Set up Julia for automatic type conversion -----------------------------------
julia <- julia_setup()
```


# Numbers  

Numbers are used to:  
+ Count objects. For example, I have 1 :apple:.    
+ Describe object behaviour. For example, the area of a :white_circle: is $\pi r^2$.     
+ Solve problems, often by introducing simple assumptions. For example, if 5 :bee: can pollinate a 100 $m^2$ field in 10 minutes, how many :bee: are required to pollinate a 350 $m^2$ field? Assume that pollination efficiency scales perfectly and is not impacted by land shape.   
+ Represent complex abstractions. For example, how can individual words be represented in an n-dimensional feature space?        

Numbers can be classified into different categories according to their properties. A number can belong to multiple categories. For example, the number 2 is a count (a natural number) and a whole number (an integer), and can also be represented as a fraction i.e. $\tfrac{2}{1}$ (a rational number) and a complex number i.e. $3+i^2$.     

Number classifications have an elegant hierarchical property. All natural numbers are integers. All integers are rational numbers i.e. $n=\tfrac{n}{1}$. All rational numbers are complex numbers i.e. $\tfrac{n}{1}=\tfrac{n+1}{1}+i^2$.      

```{r, echo=FALSE, results='markup', out.width="90%"}
knitr::include_graphics("../figures/numbers-categories.svg")
```

**Question:** Can you think of different ways to classify the number 2 compared to the number $\tfrac{5}{2}$?      


# Natural numbers  

Natural numbers ($\mathbb{N}$) are useful for describing the dimensions of a [feature space](https://stats.stackexchange.com/questions/46425/what-is-feature-space). For example, the Cartesian plane is an example of an $\mathbb{R}^2$ feature space where $\mathbb{N}=2$.         


# Real numbers  

Real numbers are sometimes useful for describing the domain or range of a function $f(x)$. The domain (D) represents all possible values that $x$ can take and the range (R) represents all possible values that $f(x)$ or $y$ can take.   

When $f(x)=3x^2$:   
+ The domain of $f(x)$ is $D:\,(-\infty,\infty)$.  
+ This can also be represented as $D:\,\{x\in\mathbb{R}\}$.   
+ The range of $f(x)$ is $R:\,[0,\infty)$.   
+ This can also be represented as $R:\,\{y\in\mathbb{R}\,|\,y\geq0\}$.     

```{r, results='hold', out.width="60%"}
# Plot the domain of f(x) = 3x^2 -----------------------------------------------
set.seed(111)

ggplot(data.frame(x = runif(1000, -3, 3)), aes(x)) +
  geom_function(fun = ~ 3*(.x)^2) +
  xlim(-2, 2) + 
  ylim(-1, 8) + 
  labs(title = "f(x) = 3x^2") +  
  theme_minimal() + 
  theme(panel.border = element_rect(fill = NA),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linetype = "dotted")) 
```

When $f(x)=\sqrt{3-x}$:  
+ The domain of $f(x)$ is $D:\;(-\infty,3]$.   
+ This can also be represented as $D:\;\{x\in\mathbb{R}\,|\,x\leq3\}$.   
+ The range of $f(x)$ is $R:\; [0,\infty)$.  
+ This can also be represented as $R:\,\{y\in\mathbb{R}\,|\,y\geq0\}$.  

```{r, results='hold', out.width="60%"}
# Plot the domain of f(x) = sqrt(3-x) ------------------------------------------
set.seed(111)

ggplot(data.frame(x = runif(1000, -3, 3)), aes(x)) +
  geom_function(fun = ~ sqrt(3-(.x))) +
  geom_vline(xintercept = 3, colour = "firebrick", linetype = "dotted") + 
  xlim(-4, 4) + 
  ylim(-1, 3) + 
  labs(title = "f(x) = sqrt(3-x)") +  
  theme_minimal() + 
  theme(panel.border = element_rect(fill = NA),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linetype = "dotted")) +
  annotate("text", x = 2.2, y = 3.6, label = "D: (-infinity, 3]")
```


# Complex numbers  

Complex numbers ($a+bi$ or $\mathbb{C}$) are represented by two components which cannot be further simplified:   
+ The term $a$ represents a real number.   
+ The term $bi$ represents the product of a real number and an imaginary number, where $i^2=-1$.   

Because the terms $a$ and $bi$ occupy completely different number planes (a real plane versus an imaginary plane), we can think of $a+bi$ as the sum of two special vectors in a special 2D space.  

The vector space for a complex number is different to the 2D Cartesian plane as its basis is $\{(1, 0), (0, i)\}$. However, the same principles of vector addition apply.   

Complex numbers therefore allow us to think about the additive and scalar multiplicative properties of vectors in $\mathbb{R}^2$. For example, the multiplication of a complex number by a scalar $k$ also scales its position in space by $k$ times.    

```{r, echo=FALSE, results='markup', out.width="90%"}
knitr::include_graphics("../figures/numbers-complex.svg")
```


# Resources  

+ A great [Wikipedia](https://en.wikipedia.org/wiki/Number#Main_classification) explanation of the number system.    
+ Introduction to number systems and binary numbers from [Khan Academy](https://www.youtube.com/watch?v=ku4KOFQ-bB4).   
+ Introduction to rational and irrational numbers from [Khan Academy](https://www.youtube.com/watch?v=cLP7INqs3JM).  
+ Numbers classification from [Nerdstudy](https://www.youtube.com/watch?v=vbPUS-0Wbv4).   