---
title: Probability theory - introduction 
author: Erika Duan  
date: "`r Sys.Date()`"  
output: 
  github_document:
    toc: true
    toc_depth: 2
    math_method:
      engine: webtex
      url: https://latex.codecogs.com/svg.image?
---  

```{r setup, include=FALSE}
# Set up global environment configuration --------------------------------------
knitr::opts_chunk$set(echo=TRUE, results='hide', fig.align='center')   
knitr::knit_engines$set(python = reticulate::eng_python)   
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}  
# Load required R packages -----------------------------------------------------
if (!require("pacman")) install.packages("pacman") 
p_load(tidyverse,
       scales,
       reticulate,
       knitr) 

# Load Anaconda environment with Python 3.9 ------------------------------------
use_condaenv("python_3_9")
```


# Introduction to probability  

Probability is an abstract concept. The frequentist approach considers probability as the relative frequency of a specific outcome. The Bayesian approach considers probability as a subjective term composed of a prior hypothesis and the observed evidence.  

However, probability can be broadly thought of as the size of a mathematical set (which can be represented in 2D as a proportion of total space).  


## Set properties  

### Scenario 1  

Imagine that we simultaneously roll two fair dice. What is the probability that the sum of two dice equals 5?   

+ We would first need to calculate all possible combinations of the dice rolls. The total number of possible combinations is the **sample space** i.e. the set of all possible outcomes. As there are 6 faces on one dice and we are rolling two dice, the sample space is $6\times6$ or $6^2$.    





# Resources  
+ The [Probability for Data Science textbook](https://probability4datascience.com/) by Stanley H Chan, specifically [Chapter 2](https://drive.google.com/file/d/1v9jLsbwG5Tl5d7XfLCfmhHuOkZZUOVNa/view) on probability     
+ Introduction to probability theory [GitHub resource](https://betanalpha.github.io/assets/case_studies/probability_theory.html) by Michael Betancourt   
+ Introduction to probability theory [Youtube series](https://www.youtube.com/playlist?list=PLUl4u3cNGP60hI9ATjSFgLZpbNJ7myAg6) from MIT   
  
