---
title: Probability theory - conditional probability
author: Erika Duan
date: "`r Sys.Date()`"
output:
  github_document:
    html_preview: FALSE
    toc: true
    toc_depth: 2
    math_method:
      engine: webtex
      url: https://latex.codecogs.com/svg.format?
---  

```{r setup, include=FALSE}
# Set up global environment configuration --------------------------------------
knitr::opts_chunk$set(echo=TRUE, results='hide', fig.align='center')
knitr::knit_engines$set(python = reticulate::eng_python)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Load required R packages -----------------------------------------------------
if (!require("pacman")) install.packages("pacman")
p_load(tidyverse,
       RColorBrewer,
       scales,
       reticulate,
       knitr)

# Load Anaconda environment with Python 3.9 ------------------------------------
use_condaenv("python_3_9")
```


# Independent events       

We previously learnt that two events can have overlapping elements or be mutually exclusive (also known as being disjointed). An example of two events being mutually exclusive is the observation that a coin can only land on either heads or tails when tossed. Mutual exclusivity therefore refers to the observation that only one or another distinct event is possible in a single outcome, which is different to the concept of independence.   

When events are independent of each other, we refer to the observation that two or more events have occurred, where the occurrence of one event does not affect the occurrence of another event (also described as where the probability of one event occurring does not change the probability of another event occurring). An example of independence is the real world knowledge that the outcome of two coin tosses are independent of each other. The observation that the first coin lands on heads does not increase the probability of the second coin also landing on heads, because each coin toss has a set probability of landing on either heads or tails.     

```{r, echo=FALSE, results='markup', out.width="70%"}
knitr::include_graphics("../figures/probability-conditional_probability-independence.svg")
```


# Conditional probability  

The opposite of independent events is conditional probability. The conditional probability of rain when the weather is cloudy or sunny is denoted as $P(rain|cloudy)$ or $P(rain|sunny)$.   

```{r}

```



# Acknowledgements  

The source materials for this tutorial are:   

+ [StackExchange answer](https://math.stackexchange.com/questions/941150/what-is-the-difference-between-independent-and-mutually-exclusive-events) on the difference between mutually exclusive and independent events    
+ Introductory Statistics [online textbook](https://saylordotorg.github.io/text_introductory-statistics/s07-03-conditional-probability-and-in.html) by Saylor Academy     
+ Independence and conditional probability [Youtube series](https://www.khanacademy.org/math/ap-statistics/probability-ap/stats-conditional-probability/v/testing-independence-from-experimental-data) from Khan academy   
+ The [Probability for Data Science textbook](https://probability4datascience.com/) by Stanley H Chan, specifically [Chapter 2](https://drive.google.com/file/d/1v9jLsbwG5Tl5d7XfLCfmhHuOkZZUOVNa/view) on probability     
+ Introduction to probability theory [GitHub resource](https://betanalpha.github.io/assets/case_studies/probability_theory.html) by Michael Betancourt   
+ Introduction to probability theory [Youtube series](https://www.youtube.com/playlist?list=PLUl4u3cNGP60hI9ATjSFgLZpbNJ7myAg6) from MIT   